Palomar 5 Gaps

### 2024-10-02 paper/main.tex
proof read the methods

    During the integration, intermediate snapshots were saved to facilitate the analysis of stellar streams and the effects of cluster impacts. Specifically, for each of the 50 realizations of the Palomar 5 stream, we saved $5000$ intermediate timesteps which is a temporal resolution of 1 million years. To provide an estimate of the size of our simulations, we have $N_p=100 000$ number of particles, $N_\textrm{time stamps} = 5 000$ time steps, $N_\textrm{phase}=6$ phase space coordinates, $N_\textrm{sampling}=50$ monte-carlo samplings, and using single precision floating point numbers, $\textrm{Single}=4$~bytes. The product of the aforementioned terms results in approximately: $N_p=100 000 \times N_\textrm{time stamps}\times N_\textrm{phase}\times N_\textrm{sampling} 4 \textrm{bytes}\approx$ 600 GB of data.

  \subsection{Gap detection and Characterization} \label{sec:gap_detection}

@@

    To facilitate this process, we ran a series of \texttt{reference} simulations. In these control simulations, we used the same 50 sets of initial conditions but excluded interactions between globular clusters and omitted the gravitational influence of the clusters on the star particles within the stream--save for the host cluster Palomar 5. By comparing the morphological differences between these control simulations and the \texttt{full} simulations (which included cluster interactions), we could identify and assess the impact of globular clusters on the stream structure more effectively.

@@

    More details on how the gaps and corresponding perturbing globular clusters were identified is provided in Appendix.~XXX. A first order approach of understanding the significance of an impact one needs to compute its change in momentum, which is determined by integrating the force acting on the particle over the duration of the interaction. A useful approximation for this change in momentum is the force at the closest approach multiplied by an estimate of the interaction time:
    \begin{equation} 
      \Delta v \approx \text{Force} \times \text{interaction time} = \frac{GM}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, 
      \end{equation} \label{eq:change_in_momentum}
    where $M$ is the mass of the perturber, $m$ is the mass of the particle, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. We compute this for the top five cloestest approaches of all globular clusters on an orbital segment about Palomar 5 that approximates the stream through out each of the 50 simulations. In the end, we have a sample of 250 impacts and compare the impact geometry of those that cause gaps to those that do not.

@@

    Although these interactions were included, we observed no significant exchanges of orbital energy or momentum between clusters.

    This experiment follows a monte-carlo scheme, where we sample the initial conditions of the globular clusters' masses and kinematics based on their corresponding uncertainties and then repeat the experiment multiple times. We choose to perform the simulation 50 times to balance computational cost with a high enough number to allow us to explore a range of possible outcomes. The uncertainties on the globular clusters are large enough that over the course of the 5 Gyr integration period, they could be anywhere allowed in their phase space given their initial energy and angular momenta. We inspected all simulations upon their completion and present one of them in in Fig.~\ref{fig:stream_on_sky}--which we selected for its clear and prominent gaps. Two of these gaps are visible in galactic coordinates and become even more apparent when marginalizing over latitutde to view it's 1D profile density over longitude.

    For each completed simulation, we follow the scheme laid out in the methods. I.e. we simulate the creation of the Palomar 5 stream for 50 different samplings of the initial conditions, with two different cases. The \textit{Full} simulation that includes the influence of all globualr clusters on the star particles representing Palomar 5 and its stream, as well as the \textit{vanilla} simulations in which the star-particles feel only the galaxy and Palomar 5. We do this for we attempt to quantify the number of gaps from globular clusters in the limit of maximum information. I.e. we are not limited by contamination of field stars, and since we are computing our own simulations, we know what a gapless stream should look like.

    This comparative analysis is shown in Fig.~\ref{fig:profiles}. We visually compare the 2D density map in tail coordinates as well as their marginalized 1D density profiles along the length of Palomar 5's orbit. As described in section~\ref{sec:gap_detection}, we locate regions on the stream where the full simulations are significantly less dense than the vanilla simulations. We find that many streams as visible by eye. However, some subtle under densities only became apparent when comparing the 1D density profiles. All simulations and labeled gaps are presented in the Appendix~\ref{appendix:gallery_of_gaps}.

    Subsequently, we followed the analysis presented in section~\ref{sec:SuspectToCulprit} to identify which globualr clusters were responsible for the gaps. This method used the orbit of Palomar 5 as a proxy for the stream. The gravitational force from all other globular clusters as computed at segments of the orbit about the position of Palomar 5 for the entire simulation. The top 5 strongest flybys were recorded to identify which cluster was responsible. This is shown in Fig.~\ref{fig:force-on-orbit}, where the top panel shows the force from the globular clusters on the orbit, while the bottom panel shows the density of the stream. A density threshold was imparted to quantify the length of the stream.  The location and time of impact was cross referenced to the location of the gaps to identify which cluster was responsible.

@@

@@

  \section{Gap detection and characterization methods}
    \subsubsection*{Density Maps and Profiles}
      For each of the 50 simulations, we compare the final snapshots between the \texttt{vanilla} and \texttt{full} simulations using the tail coordinate system. We inspect these differences by generating 2D density maps of the streams. Additionally, we marginalize over the $y'$-coordinate to produce 1D density profiles along the $x'$-axis. These comparisons are shown in the results section (see Fig.~\ref{fig:profiles}).

@@

    \subsubsection*{From Suspect to Culprit} \label{sec:SuspectToCulprit}

@@

    \subsection{Reconstruction of the impact geometry}

@@

      To achieve this, we begin by identifying the approximate moments of impact from the most significant clusters, as determined in the previous analysis in Sec.~\ref{sec:SuspectToCulprit}. Then, we refine these estimates. While the orbit serves as a good first-order proxy for the stream, it is insufficient, as the typical offset between the stream and its orbit can be on the order of 200-400 pc—--significantly larger than the impact parameters we are investigating.

>>>>

    During the integration, intermediate snapshots were saved to facilitate the analysis of stellar streams and the effects of cluster impacts. Specifically, for each of the 50 realizations of the Palomar 5 stream, we saved $5000$ intermediate timesteps which is a temporal resolution of 1 million years. To provide an estimate of the size of our simulations, we have $N_p=100 000$ particles, $N_\textrm{time stamps} = 5 000$ time steps, $N_\textrm{phase}=6$ phase space coordinates, $N_\textrm{sampling}=50$ monte-carlo samplings, and using single precision floating point numbers, $\textrm{Single}=4$~bytes. The product of the aforementioned terms results in approximately: $N_p \times N_{\textrm{time stamps}}\times N_{\textrm{phase}}\times N_{\textrm{sampling}} \times 4 \textrm{bytes}\approx$ 600 GB of data.

  \subsection{Gap Detection and Characterization} \label{sec:gap_methods}

@@

    To facilitate this process, we conducted a series of \texttt{reference} simulations. These control simulations used the same 50 sets of initial conditions but excluded interactions with globular clusters, except for the host cluster, Palomar 5. In other words, the gravitational influence of all other clusters on the star particles within the stream was omitted. By comparing the morphological differences between these control simulations and the \texttt{full} simulations (which included cluster interactions), we could more effectively identify and assess the impact of globular clusters on the stream's structure. Details on gap identification are provided in Appendix~\ref{sec:gap_detection}, and the identification of corresponding perturbing globular clusters is detailed in Appendix~\ref{sec:Perturber_Identification}.

@@

    Once the gaps and perturbers were identified, we sought to characterize the properties of the globular clusters that caused gaps and differentiate them from those that did not. Carlberg (2012) initiated the theory of gap creation in stellar streams due to gravitational flybys. This analysis was extended both analytically and numerically by Erkal et al. (2015), who studied uniform-density streams on circular orbits. Sanders et al. (2016) presented the most comprehensive theory, generalizing Erkal et al.'s work to include impacts on streams of varying densities on inclined and eccentric orbits using an action-angle approach.

    The significance of an impact is measured by the change in momentum imparted to the stream stars by the perturber. To first order, we follow the developments from collision theory as laid out in Galactic Dynamics (Binney \& Tremaine) and similar works by John Taylor. In the simplest case, the perturber is considered unperturbed during the interaction, and thus has constant velocity. We can consider the most basic scenario of one point mass impacting another. A useful approximation for the change in momentum is the force at the closest approach multiplied by an estimate of the interaction time: \begin{equation} \Delta v \approx \text{Force} \times \text{interaction time} = \frac{GM}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, \end{equation} where $M$ is the mass of the perturber, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. We compute this for the five closest approaches of all globular clusters throughout each of the 50 simulations. Details on how this computation was performed are given in Appendix~\ref{sec:reconstruction}. Ultimately, we obtained a sample of 250 impacts and compared the impact geometry of those that caused gaps with those that did not.

@@

    Firstly, to our knowledge this is the first time any one tried to study the interactions between the globular clusters. By including N~body forces on the orbits with the globular clusters, we did not observe any significant changes to orbital properties such as energy or angular momentum between any of the globular clusters in any of the simulations. This is an expected result given their small sizes and vast distances between them.

   We inspected all simulations upon their completion and present one of them in in Fig.~\ref{fig:stream_on_sky}--which we selected for its clear and prominent gaps. Two of these gaps are visible in galactic coordinates and become even more apparent when marginalizing over latitutde to view it's 1D profile density over longitude.

    For each completed simulation, we follow the scheme laid out in the methods. I.e. we simulate the creation of the Palomar 5 stream for 50 different samplings of the initial conditions, with two different cases. The \textit{Full} simulation that includes the influence of all globualr clusters on the star particles representing Palomar 5 and its stream, as well as the \textit{vanilla} simulations in which the star-particles feel only the galaxy and Palomar 5.

    This comparative analysis is shown in Fig.~\ref{fig:profiles}. We visually compare the 2D density map in tail coordinates as well as their marginalized 1D density profiles along the length of Palomar 5's orbit. As described in section~XXX, we locate regions on the stream where the full simulations are significantly less dense than the vanilla simulations. We find that many streams as visible by eye. However, some subtle under densities only became apparent when comparing the 1D density profiles. All simulations and labeled gaps are presented in the Appendix~\ref{appendix:gallery_of_gaps}.

    Subsequently, we followed the analysis presented in section~\ref{sec:Perturber_Identification} to identify which globualr clusters were responsible for the gaps. This method used the orbit of Palomar 5 as a proxy for the stream. The gravitational force from all other globular clusters as computed at segments of the orbit about the position of Palomar 5 for the entire simulation. The top 5 strongest flybys were recorded to identify which cluster was responsible. This is shown in Fig.~\ref{fig:force-on-orbit}, where the top panel shows the force from the globular clusters on the orbit, while the bottom panel shows the density of the stream. A density threshold was imparted to quantify the length of the stream.  The location and time of impact was cross referenced to the location of the gaps to identify which cluster was responsible.

@@

    We do this for we attempt to quantify the number of gaps from globular clusters in the limit of maximum information. I.e. we are not limited by contamination of field stars, and since we are computing our own simulations, we know what a gapless stream should look like.

@@

  \section{Gap Detection} \label{sec:gap_detection}
    For each of the 50 simulations, we compare the final snapshots between the \texttt{vanilla} and \texttt{full} simulations using the tail coordinate system. We inspect these differences by generating 2D density maps of the streams. Additionally, we marginalize over the $y'$-coordinate to produce 1D density profiles along the $x'$-axis. These comparisons are shown in the results section (see Fig.~\ref{fig:profiles}).

@@

    \section{Perturber Identification} \label{sec:Perturber_Identification}

@@

    \section{Reconstruction of the impact geometry} \label{sec:reconstruction}

@@

      To achieve this, we begin by identifying the approximate moments of impact from the most significant clusters, as determined in the previous analysis in Sec.~\ref{sec:Perturber_Identification}. Then, we refine these estimates. While the orbit serves as a good first-order proxy for the stream, it is insufficient, as the typical offset between the stream and its orbit can be on the order of 200-400 pc—--significantly larger than the impact parameters we are investigating.
